---
title: "Co-registration for the CRC LM55 image"
author: "Sai Srikanth Lakkimsetty"
date: "Aug, 2022"
output:
    BiocStyle::html_document:
    toc: true 
vignette: >
  %\VignetteIndexEntry{Co-registration for the CRC dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
# A flag to manage chunk re-runs while knitting ... 
# Set SHARING to FALSE if you are re-knitting the code and variables are already 
# in the environment. Manage individual code chunk evaluation with "eval" option. 
SHARING <- FALSE 

if (SHARING) {
    library("msireg") 
} else {
    library("devtools")
    load_all()
} 

library("ggplot2")
RNGkind("L'Ecuyer-CMRG") 

setCardinalBPPARAM(MulticoreParam(stop.on.error=FALSE, progressbar=TRUE))
setCardinalVerbose(TRUE)
```

```{r style, echo=FALSE, results='asis', eval=TRUE}
BiocStyle::markdown()
```


# Introduction  

In this vignette, we illustrate the co-registration between mass spectrometry imaging of a colorectal cancer tumor tissue of liver matastases and H&E stained microscopic image. The mass spectrometry image is collected using a 4800 MALDI-TOF/TOF Analyzer at a step size of 150µm. The dataset is available via [PRIDE](https://www.ebi.ac.uk/pride/archive/projects/PXD039409). More information on the data acquisition methodology can be found in this [preprint manuscript](https://www.preprints.org/manuscript/202302.0363/v1). 

There are four key inputs to the registration workflow. These are the fixed image, the moving image, MSI region of interest (ROI) and optical image ROI. We consider the MS image as the *fixed image* and the optical image as the *moving image*. That means, the optical (microscopic) image is transformed to align with the MS image. The regions of interest (ROIs) provide the registration method with information on where the tissue is present in the slide. It is essential that the registration method knows where the background is present to exclude it from registration process. 

Similarly there are four key steps to registering MS and optical images. We shall discuss more about the steps as we perform them. 

* Pre-processing MS and optical images 
* ROI selection of the images
* Pre-alignment 
* Registration 

We load the MS and optical images into R as below. 

```{r data-load, eval=SHARING} 
# DATAPATH <- "/Users/lakkimsetty.s/Documents/9-data"
# CURRPATH <- "9-CRC"
# 
# opt_filename <- "b_prim_LM55- HE.png"
# opt <- EBImage::readImage(file.path(DATAPATH, CURRPATH, opt_filename))
# opt <- flip(opt) 
# opt <- resize(opt, w=dim(opt)[1]/2, h=dim(opt)[2]/2)
# 
# mse_filename <- "LM55"
# mse <- readAnalyze(name=mse_filename,
#     folder=file.path(DATAPATH, CURRPATH))
```


```{r data-load-2, eval=SHARING} 
DATAPATH <- "/Users/lakkimsetty.s/Documents/9-data"
CURRPATH <- "9-CRC"
load(file.path(DATAPATH, CURRPATH, "processed-in-R/mse-peaks-LM55.RData"))
```


# MS image pre-processing 

Before pre-processing the MSI data, we specify the a parallelization-safe RNG to ensure reproducibility even on parallel computation. Pre-processing is performed to clean up the spectra by removing noise. It includes peak picking, aligning, filtering and binning. We first construct a mean spectrum across all spectra. 

```{r pre-proc, eval=SHARING}  
mse_mean <- summarizeFeatures(mse, FUN="mean",
    BPPARAM=MulticoreParam(workers=multicoreWorkers()))

ref <- mse_mean |>
    peakPick(method="mad", SNR=2) |>
    peakAlign() |>
    peakFilter() |>
    process()

mse_peaks <- mse |>
    normalize(method="tic") |>
    peakBin(ref=mz(ref), tolerance=100, units="ppm") |>
    process()
``` 


Once the pre-processing is completed, we save the MSI data (peaks) to avoid having to pre-process the data again in case we want to re-run the registration workflow. 

```{r save-proc-imzml, eval=SHARING} 
# writeMSIData(mse_peaks,
#              file.path(DATAPATH, CURRPATH,
#                        "processed-in-R/HR2MSI-mouse-bladder-peaks.imzML"))
```


# Identifying regions-of-interest (ROIs) 

First, we make a list of the processed data attributes: X and Y spatial resolution of processed MSI data; number of features and pixels of processed MSI data; and X and Y spatial resolution of optical image. 

```{r mse-params, eval=SHARING} 
mse_attrs <- list(
    nX = dims(mse_peaks)[1], nY = dims(mse_peaks)[2],
    nF = dim(mse_peaks)[1], nP = dim(mse_peaks)[2],
    nXo = dim(opt)[1], nYo = dim(opt)[2]
)
```


## MSI ROI 
`msireg` provides an in-built tool to generate ROI for MSI tissue. We select ROI using the function `multiSelectROI` which is built on `Cardinal`'s `selectROI` method. `multiSelectROI` can select multiple disconnected tissue regions in one go. The method `multiSelectROI` needs an `mz` value to display a raster on which ROI is selected. 

If you already have a `mz` value that properly demarcates tissue and background, skip the next two code blocks. Otherwise, we build a quick spatial shrunken centroid model (refer `?spatialShrunkenCentroids` for details) to come up with a list of `mz` values that are suitable for this task. The `mz` list is derived from the `ssc` model's top features. This is a reliable way to select one `mz` value for ROI selection. 

```{r viz-ssc, eval=SHARING} 
set.seed(2)
ssc <- spatialShrunkenCentroids(mse_peaks, r=c(1), s=c(24), k=c(15), 
    method="gaussian", BPPARAM=MulticoreParam()) 
topf <- topFeatures(ssc, n=10, model=list(s=24))$mz 
``` 


Now we select a `mz` value from the top features`topf`. `mz = 502.978` is a good ion image to draw MSI ROI. 

```{r viz-ssc-select-mz}
image(mse_peaks, mz=topf[2]) 
message("Appropriate m/z value for ROI selection: ", round(topf[2], 3))
```


To select the ROI using a mouse on the plotting pane, run the below command directly from console. We specify the `mz` value we selected above. In fact, `multiSelectROI` can be used on a variety of plotting types including superposed image plots. 

```{r msi-roi, eval=SHARING} 
# RUN DIRECTLY FROM CONSOLE, SESSION WILL CRASH OTHERWISE 
mse_roi <- multiSelectROI(mse_peaks, mz=502.978) 
``` 


## Optical image ROI 

We use `multiDrawROI` to select ROI for optical images. The usage is identical to that of `multiSelectROI` (without the need for a `mz` value). It is to be executed directly from console to prevent R session from crashing. 

```{r opt-roi, eval=SHARING}
# RUN DIRECTLY FROM CONSOLE, SESSION WILL CRASH OTHERWISE 
opt_roi <- multiDrawROI(opt)
```


# Pre-alignment of images

So far, we have performed pre-processing and ROI selection of the images. Before doing the registration step, we process the images to ensure the images are overlapping. Often, the tissues on the imaged slide are at different (non-overlapping) positions. That means the burden of first bringing the images together falls on the registration method, which could affect registration performance. To alleviate this problem, we ensure overlap of two images by cropping and padding. We crop and/or pad optical (moving) image so that it overlaps with the tissue of the MS (fixed) image. To clarify, we are not registering the images in this step (no transformations of the grid) but we do keep track of the the cropping and padding operations to ensure reproducibility. 

```{r opt-mask-img-crop, eval=SHARING} 
OUT <- cropToEdgesAndPadZeros(ref_mask=mse_roi, target_mask=opt_roi, target_img=opt) 
opt <- OUT$img 
opt_roi <- OUT$mask 
``` 


# Registration 

We register the MS image (pre-processed `mse_peaks`) and the optical image (`opt`) using the `coregister`. A scaling factor, `spatial_scale`, allows the method to register the images at a higher / lower resolution (with respect to the `mse`'s spatial dimensions). This is to overcome the imaging resolution disparity between MSI and optical images. The method returns a list. The registration components can be accessed using the `$`. 

```{r coreg, eval=SHARING}
OUT <- coregister(mse_peaks, opt, mse_roi, opt_roi, spatial_scale=4, 
                  BPPARAM=MulticoreParam(), verbose=TRUE)
```


```{r, coreg-write-to-file, eval=SHARING} 
save.image("~/Documents/1-saved-envs/vignettes/7-3-CRC-LM56.RData") 
print(getwd())
WriteTransform(OUT$reg$TF, "../../../data/saved-transforms/7-3-CRC-LM56.hdf5")
```


We visualize the results of the registration. We use the method `Resample` from `SimpleITK` to obtain the transformed moving image. The plot below shows the overlays of the MS (fixed) image and the optical (moving) images' overlays before and after registration. 

```{r viz-coreg-prep, eval=TRUE} 
# library("devtools")
# load_all()
# load("~/Documents/1-saved-envs/vignettes/7-4-CRC-LM55.RData")
# OUT$reg$TF <- ReadTransform("data/saved-transforms/7-4-CRC-LM55.hdf5")
# out <- rlang::duplicate(OUT)
# out$fixed <- SimpleITK::as.image(out$opt, isVector=FALSE)
# out$moving <- SimpleITK::as.image(out$msimg, isVector=FALSE)

out$msimgx <- Resample(out$moving, out$reg$TF) |> as.array() |> Image() 

overlay <- overlayGridOnImage(out$msimgx)$overlay |> as.image()
grid <- overlayGridOnImage(out$msimgx)$grid |> as.image()

out$overlayx <- Resample(overlay, out$reg$TF) |> as.array() |> Image()
out$gridx <- Resample(grid, out$reg$TF) |> as.array() |> Image()
```


```{r viz-coreg} 
alpha <- 0.45
out$movingx <- Resample(out$moving, out$reg$TF)
.fixed <- out$fixed |> as.array() |> Image()
.moving <- out$moving |> as.array() |> Image()
.movingx <- out$movingx |> as.array() |> Image()

ol1 <- (.fixed * alpha + .moving * (1-alpha))
ol2 <- (.fixed * alpha + .movingx * (1-alpha))

def.par <- par(no.readonly=TRUE)

height <- 7
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

titles <- c("Optical image", "Mass spec. image",
            "Overlay before registration", "Overlay after registration")

par(mar=c(0.25, 0.25, 0.75, 0.25))
pl <- layout(mat=matrix(c(1:4), ncol=2, byrow=TRUE),
             heights=rep(height/2, 2), widths=rep(width/2, 2),
             respect=TRUE)

. <- sapply(c(1:4), function(x) {
    plot.new()
    rasterImage(list(.fixed, .moving, ol1, ol2)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE)
    title(titles[x])
})
```


```{r coreg-grid-overlay-viz} 
height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("warping on grid", "overlay with grid warping") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(out$gridx, out$overlayx)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


```{r metric-dice} 
spatial_scale <- 4 
DIM <- dim(out$MSIMG)[1:2] * spatial_scale
D_prereg <- diceCoefFromMasks(mse_roi, opt_roi) 
mse_roi_ss <- resizeAndPadImageToMatchDims(mse_roi, DIM) 
opt_roi_ss <- resizeAndPadImageToMatchDims(opt_roi, DIM) 
mse_roi_ss_x <- applyTransformOnMask(mse_roi_ss, out$reg$TF) 
D_postreg <- diceCoefFromMasks(mse_roi_ss_x, opt_roi_ss) 
```

Dice coefficient is computed from MSI and OPT masks. Dice coefficient before 
and after registration is `r D_prereg` and `r D_postreg` respectively. 

```{r metric-dice-viz} 
ol3 <- t(opt_roi_ss * 0.5 + mse_roi_ss * 0.5)
ol4 <- t(opt_roi_ss * 0.5 + mse_roi_ss_x * 0.5)

height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("Masks overlap before coreg", "Masks overlap after coreg") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(ol3, ol4)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


```{r ncc} 
ncc_prereg <- normalizedCrossCorr(.fixed, .moving) 
ncc_postreg <- normalizedCrossCorr(.fixed, .movingx) 
```

Normalized cross correlation is a similarity metric that measure correlation 
between pixel intensities. It ranges from `-1` to `1`. The values before and 
after co-registration are `r ncc_prereg` and `r ncc_postreg` respectively. 


```{r jacobian} 
jac <- jacobianDeterminantFromTransform(out$reg$TF, ref_img=out$moving) 
hist(jac$jacobian_det, breaks=40)
```


```{r jacobian-viz-try1, eval=FALSE} 
jac$jacobian_det |> EBImage::display("raster") 
imagePlot(t(flip(jac$jacobian_det)))
```

```{r, jacobian-viz, eval=TRUE} 
def.par <- par(no.readonly=TRUE)
jacdet <- jac$jacobian_det
d <- dim(jacdet) 
asp <- d[2] / d[1] # h / w 
w <- 9 
h <- asp * w 

jacdet[jacdet <= 0] <- 0 
nfolds <- sum((jacdet == 0))
nshrinks <- length(unique(jacdet[(jacdet > 0) & (jacdet < 1)])) 
nequals <- sum(jacdet == 1)
nexpands <- length(unique(jacdet[(jacdet > 1)])) 

folds <- if (nfolds == 0) c() else c("#FF0000") # red 
shrinks <- hcl.colors(n=nshrinks, palette="Oranges") # white to oranges
equals <- if (nequals == 0) c() else c("#FFFFFF") # white
expands <- hcl.colors(n=nexpands, palette="Blues", rev=TRUE) # white to blues

layout(mat=matrix(1, nrow=1, ncol=1), heights=h, widths=w, respect=TRUE)
fields::imagePlot(flip(jacdet), asp=d[2]/d[1], xaxt="n", yaxt="n", 
          col=c(folds, shrinks, equals, expands), 
          lowerTriangle=TRUE)

par(def.par)
```


```{r, eval=FALSE} 
library("scales")
mesh <- expand.grid(x=c(1:d[1]), y=c(1:d[2])) 
mesh$z <- c(flip(jacdet)) 

ggplot(data=mesh, aes(x=x, y=y)) + 
    geom_tile(aes(fill=z)) + 
    scale_fill_gradient2(low = "red", 
                         mid = "white", 
                         high = "blue", 
                         midpoint=1) + 
    theme_void()
```


```{r, eval=FALSE}
def.par <- par(no.readonly=TRUE)
jacdet <- jac$jacobian_det
d <- dim(jacdet) 
asp <- d[2] / d[1] # h / w 
w <- 9 
h <- asp * w 

ncolors <- length(unique(jacdet))
layout(mat=matrix(1, nrow=1, ncol=1), heights=h, widths=w, respect=TRUE) 

x <- c(1:d[1]) 
y <- c(1:d[2])
image(x, y, (flip(jacdet)), zlim=range(jacdet), xlim=range(x), ylim=range(y), 
      xlab="", ylab="", xaxt="n", yaxt="n", 
      col=hcl.colors(ncolors))
```


The range of the determinant of the Jacobian is `r range(jac$jacobian_det)`. The 
histogram shows that most of the det values are around 1. That's good. The 
number of pixel locations where folding has occurred is 
`r sum(jac$jacobian_det <= 0)`.
Values 

* = 1 imply no area (or volume) change 
* (> 1) imply increased area
* (1, 0) imply reduced area 
* < 0 imply folding <- this is bad 


```{r gradient-field-cosine-dist, eval=SHARING} 
img1 <- .fixed
img2 <- .moving
img2x <- .movingx

grads1 <- matrixGradients(imageData(img1))
grads2 <- matrixGradients(imageData(img2))
grads2x <- matrixGradients(imageData(img2x))

.grads1 <- normalizeGradients(grads1)
.grads2 <- normalizeGradients(grads2)
.grads2x <- normalizeGradients(grads2x)

D <- dim(img1)

.grads <- array(c(.grads1, .grads2), dim=c(D, 4))
cos_dist <- apply(.grads, c(1,2), function(x) {
    if (identical(x[c(1:2)], x[c(3:4)])) return(1)

    (sum( x[c(1:2)] * x[c(3:4)] ))
}) |>
    array(dim=dim(img1))


.gradsx <- array(c(.grads1, .grads2x), dim=c(D, 4))
cos_distx <- apply(.gradsx, c(1,2), function(x) {
    if (identical(x[c(1:2)], x[c(3:4)])) return(1)

    (sum( x[c(1:2)] * x[c(3:4)] ))
}) |>
    array(dim=dim(img1))
```


```{r gradient-cosine-dist-plots, eval=TRUE} 
height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("Cosine dist before reg", "Cosine distance after reg") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(t(1-cos_dist), t(1-cos_distx))[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


# Session info

```{r session-info} 
sessionInfo()
```
