---
title: "Installation instructions for MSIreg"
author: "Sai Srikanth Lakkimsetty, Andreas Weber, Dr. Niko Pinter" 
date: "Apr, 2024"
output:
    BiocStyle::html_document: 
    html_document: 
        toc: true 
        toc_float: true
        toc_collapsed: true
vignette: >
  %\VignetteIndexEntry{Installation instructions for MSIreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---


```{r style, echo=FALSE, results="asis", eval=TRUE}
BiocStyle::markdown() 
```


# Introduction 

`MSIreg` is an open-source R package for coregistering mass spectrometry images with H&E images. Unfortunately, due to the dependency `SimpleITK`, the package could not be directly installed from CRAN or Bioconductor. However, using the `remotes` package, `MSIreg` can be installed on MacOS and Linux based operating systems directly through `R`. 

Installing software packages on Windows operating system can sometimes be a challenge due to compatibility issues. However, with the help of Windows Subsystem for Linux (WSL) and RStudio Server, users can overcome these obstacles and install packages such as `MSIreg` with ease.

In this guide, we will walk you through the steps to install `MSIreg` on MacOS, Linux using the `remotes` package and on Windows using WSL and RStudio Server. 


# MacOS and Linux based operating systems 

Before installing `MSIreg`, we will install `Cardinal` through the Bioconductor repository and then `SimpleITK` though GitHub with the `remotes` package. 


```{r, eval=FALSE} 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Cardinal")  
```

```{r, eval=FALSE} 
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("SimpleITK/SimpleITKRInstaller") 
``` 

Now that the dependencies have been installed, we will install `MSIreg` through GitHub. 

```{r, eval=FALSE} 
remotes::install_github("sslakkimsetty/msireg") 
```


# Windows operating system 

The Windows OS and the computer should support the usage of Windows Subsystem for Linux (WSL) for this procedure to succeed. 

## Create Conda environment

Open an Ubuntu terminal shell and create a new Conda environment called `msireg` by running the following command. 

```{r, engine="bash", eval=FALSE}
conda create -n msireg
```


## Activate Conda environment

Activate the newly created Conda environment by running the following command. 

```{r, engine="bash", eval=FALSE}
conda activate msireg
```

## Install necessary packages

Next, install the necessary packages using the following command from within the Conda environment. 
```{r, engine="bash", eval=FALSE}
conda install -c conda-forge -c bioconda r-base r-devtools r-remotes SimpleITK bioconductor-ebimage bioconductor-cardinal bioconductor-biocparallel r-rtsne
```

Consider adding these version numbers to avoid dependency incompatibilities with MSIreg

```{r, engine = 'bash', eval = FALSE}
conda install -c conda-forge -c bioconda r-base=4.2.2 r-devtools=2.4.5 r-remotes SimpleITK=2.2.1 bioconductor-ebimage=4.40.0 bioconductor-cardinal=3.0.1 bioconductor-biocparallel=1.32.5 r-rtsne=0.16
```


## Install `MSIreg` package

Open `R` by running the following command in the terminal from within the Conda environment. 

```{r, engine="bash", eval=FALSE} 
R 
```

Once R is open, install the `MSIreg` package by running the following command within `R`:

```{r, eval=FALSE}
remotes::install_github("sslakkimsetty/msireg") 
```


## Install RStudio server

To run RStudio in the browser, first download and install RStudio server from [Posit](https://posit.co/download/rstudio-server/). Follow the instructions provided on the website for Debian/Ubuntu to complete the installation.

Run the following commands in the Ubuntu terminal. You may have to update the download link when updates to RStudio happen. 

```{r, engine="bash", eval=FALSE}
sudo apt-get install gdebi-core
wget https://download2.rstudio.org/server/jammy/amd64/rstudio-server-2023.03.0-386-amd64.deb
sudo gdebi rstudio-server-2023.03.0-386-amd64.deb
```


## Start RStudio server within the activated conda environment

Activate the Conda environment by running the following command in the Ubuntu terminal. 

```{r, engine="bash", eval=FALSE}
sudo /usr/lib/rstudio-server/bin/rserver \
--server-daemonize=0 \
--www-port 8787 \
--rsession-which-r=$(which R) \
--rsession-ld-library-path=$CONDA_PREFIX/lib
```

## Access RStudio server

Access RStudio server by opening your web browser and navigating to `http://localhost:8787` 

Log in to RStudio with your Ubuntu credentials.


# Session info 
```{r session-info, eval=TRUE} 
sessionInfo()
```

