---
title: "Installation instructions for MSIreg"
author: "Sai Srikanth Lakkimsetty, Andreas Weber, Dr. Niko Pinter" 
date: "Apr, 2024"
output:
    BiocStyle::html_document: 
    html_document: 
        toc: true 
        toc_float: true
        toc_collapsed: true
vignette: >
  %\VignetteIndexEntry{Installation instructions for MSIreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---


```{r style, echo=FALSE, results="asis", eval=TRUE}
BiocStyle::markdown() 
```


# Introduction 

`MSIreg` is an open-source R package for coregistering mass spectrometry images with H&E images. Unfortunately, due to the dependency `SimpleITK`, the package could not be directly installed from CRAN or Bioconductor. However, using the `remotes` package, `MSIreg` can be installed on MacOS and Linux based operating systems directly through `R`. 

Installing software packages on Windows operating system can sometimes be a challenge due to compatibility issues. However, with the help of Windows Subsystem for Linux (WSL) and RStudio Server, users can overcome these obstacles and install packages such as `MSIreg` with ease.

In this guide, we will walk you through the steps to install `MSIreg` on MacOS, Linux using the `remotes` package and on Windows using WSL and RStudio Server. 

We will finish with some tools and resources for converting raw MSI from instruments to imzML format. 


# MacOS and Linux based operating systems 

Before installing `MSIreg`, we will install `Cardinal` through the Bioconductor repository and then `SimpleITK` though GitHub with the `remotes` package. 


```{r, eval=FALSE} 
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Cardinal")  
```

```{r, eval=FALSE} 
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("SimpleITK/SimpleITKRInstaller") 
``` 

Now that the dependencies have been installed, we will install `MSIreg` through GitHub. 

```{r, eval=FALSE} 
remotes::install_github("sslakkimsetty/msireg") 
```


# Windows operating system 

The Windows Subsystem for Linux (WSL) is a feature of Windows OS that lets developers access both Windows and Linux machines on the same computer. You must be running Windows 10 version 2004 and higher (Build 19041 and higher) or Windows 11 to use the installation method below. If you are on earlier versions please see the [manual install page](https://learn.microsoft.com/en-us/windows/wsl/install-manual). 

An alternative WSL installation method is described in the [YouTube video](https://www.youtube.com/watch?v=av0UQy6g2FA). 

## Install WSL 

Open PowerShell or Windows Command Prompt in administrator mode by right-clicking and selecting "Run as admistrator" and execute the following command. 

```{r, engine="bash", eval=FALSE} 
wsl --install 
```


The default linux distribution installed is Ubuntu. This can be changed by using the `-d` flag. 

```{r, engine="bash", eval=FALSE} 
wsl --list --online # lists available distribution

# Select an option from above list and install 

wsl --install -d <Distribution Name> 
```

Once WSL and a linux distribution (OS) has been installed, you will need to create a user accound and password for it. Please see the best practices for setting up WSL [here](https://learn.microsoft.com/en-us/windows/wsl/setup/environment#set-up-your-linux-username-and-password). 

Once WSL is set up, it is time to create a conda environment. 

## Interoperabilty of Windows and Linux file systems 

WSL does not require additional hardware and runs entirely on the Windows machine. More specifically, WSL is a feature of Microsoft Windows that allows for running Linux programs directly within Windows without the need for a virtual machine or extra hardware. WSL also shares the same filesystem as Windows so there would be no need for file transfers or extra networking equipment. (Users can write/read files on the Windows drives directly from within the Linux subsystem.) 
For more information on working across Windows and Linux file systems and interoperability, please see [this link](https://learn.microsoft.com/en-us/windows/wsl/filesystems). 

## Create Conda environment

Open an Ubuntu terminal shell and create a new Conda environment called `msireg` by running the following command. 

```{r, engine="bash", eval=FALSE}
conda create -n msireg
```


## Activate Conda environment

Activate the newly created Conda environment by running the following command. 

```{r, engine="bash", eval=FALSE}
conda activate msireg
```

## Install necessary packages

Next, install the necessary packages using the following command from within the Conda environment. 
```{r, engine="bash", eval=FALSE}
conda install -c conda-forge -c bioconda r-base r-devtools r-remotes SimpleITK bioconductor-ebimage bioconductor-cardinal bioconductor-biocparallel r-rtsne
```

Consider adding these version numbers to avoid dependency incompatibilities with MSIreg

```{r, engine = 'bash', eval = FALSE}
conda install -c conda-forge -c bioconda r-base=4.2.2 r-devtools=2.4.5 r-remotes SimpleITK=2.2.1 bioconductor-ebimage=4.40.0 bioconductor-cardinal=3.0.1 bioconductor-biocparallel=1.32.5 r-rtsne=0.16
```


## Install `MSIreg` package

Open `R` by running the following command in the terminal from within the Conda environment. 

```{r, engine="bash", eval=FALSE} 
R 
```

Once R is open, install the `MSIreg` package by running the following command within `R`:

```{r, eval=FALSE}
remotes::install_github("sslakkimsetty/msireg") 
```


## Install RStudio server

To run RStudio in the browser, first download and install RStudio server from [Posit](https://posit.co/download/rstudio-server/). Follow the instructions provided on the website for Debian/Ubuntu to complete the installation.

Run the following commands in the Ubuntu terminal. You may have to update the download link when updates to RStudio happen. 

```{r, engine="bash", eval=FALSE}
sudo apt-get install gdebi-core
wget https://download2.rstudio.org/server/jammy/amd64/rstudio-server-2023.03.0-386-amd64.deb
sudo gdebi rstudio-server-2023.03.0-386-amd64.deb
```


## Start RStudio server within the activated conda environment

Activate the Conda environment by running the following command in the Ubuntu terminal. 

```{r, engine="bash", eval=FALSE}
sudo /usr/lib/rstudio-server/bin/rserver \
--server-daemonize=0 \
--www-port 8787 \
--rsession-which-r=$(which R) \
--rsession-ld-library-path=$CONDA_PREFIX/lib
```

## Access RStudio server

Access RStudio server by opening your web browser and navigating to `http://localhost:8787` 

Log in to RStudio with your Ubuntu credentials. 


# Conversion tools from RAW to imzML 

RAW data acquired from different vendors need different ways of converting into imzML. Below are some links and resources for different vendors. 

- Thermo: [converter for LTQ- and Orbitrap-based instruments](https://www.ms-imaging.org/imzml/software-tools/raw-to-imzml-converter/); imzML export function is implemented in the new version of [ImageQuest](https://www.thermofisher.com/order/catalog/product/10137985) software. 
- Waters: imzML export function in [High Definition Imaging](http://www.waters.com/waters/en_US/High-Definition-Imaging-%28HDI%29-Software/nav.htm?locale=en_US&cid=134833914) software. 
- Bruker: imzML export function in [flexImaging](https://www.bruker.com/products/mass-spectrometry-and-separations/ms-software/fleximaging/overview.html) and [SCiLS Lab](https://www.bruker.com/products/mass-spectrometry-and-separations/ms-software/scils/overview.html) software. 
- ABI SCIEX: use `toimzml` module 
- Shimadzu: imzML export function in [Imaging MS Solution 1.1](https://www.shimadzu.eu/imaging-mass-solution-ver11-software). 
- mzML: [imzML converter download link](http://www.imzmlconverter.co.uk/) 
- METASPACE: [link to converting high mass resolution data to imzML](http://metaspace2020.eu/imzml). 


# Session info 
```{r session-info, eval=TRUE} 
sessionInfo()
```

