---
title: "Stochasticity observation in `msireg` package"
author: "Sai Srikanth Lakkimsetty"
date: "Feb, 2023"
output:
    BiocStyle::html_document:
    toc: true 
vignette: >
  %\VignetteIndexEntry{Stochasticity observation in `msireg` package} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
# A flag to manage chunk re-runs while knitting 
# (could also use cache option, but tricky) 
# Set SHARING to FALSE if you are re-knitting the code and variables are already 
# in the environment. Manage individual code chunk evaluation with "eval" option. 
SHARING <- FALSE 

if (SHARING) {
    library("msireg") 
} else {
    library("devtools")
    load_all()
} 

RNGkind("L'Ecuyer-CMRG") 

setCardinalBPPARAM(MulticoreParam(stop.on.error=FALSE, progressbar=TRUE))
setCardinalVerbose(TRUE)
```

```{r style, echo=FALSE, results="asis", eval=TRUE}
BiocStyle::markdown()
```


# Introduction  

In this vignette, we try to reproduce the stochasticity of processed grayscaled versions of MSI images obtained through the `msireg` package. We achieve this analysis through the same series of steps that `msireg` takes. The dataset we use is the mouse bladder one. 

* First, we investigate if this is the result of `Cardinal`'s pre-processing methods. This is unlikely but we include this in the instrest of being comprehensive. 
* Second, we see if SSC is the culprit. This could be very likely as `msireg` does set a seed. 
* Lastly, we move on to `Rtsne`. 


```{r data-load, eval=SHARING} 
DATAPATH <- "/Users/lakkimsetty.s/Library/Mobile Documents/com~apple~CloudDocs/00-NEU/2-Ind-Study/9-data"
CURRPATH <- "2-hr2msi-mouse-urinary-bladder"

msi_filename <- "HR2MSI mouse urinary bladder S096.imzML"

mse <- readMSIData(file.path(DATAPATH, CURRPATH, msi_filename))
``` 


# \texttt{Cardinal}'s pre-processing 


| Performance metric | Before coreg | After coreg | 
|:-------------------|:------------:|:-----------:| 
| Dice coefficient | 95.52% | 99.59% | 
| Normalized cross correlation | 86.3% | 95.4% | 
| Jacobian determinant | - | 0.25% (folding) | 


| Performance metric | Before coreg | After coreg | 
|:-------------------|:------------:|:-----------:| 
| Dice coefficient | 82.66% | 95.74% | 
| Normalized cross correlation | 75.6% | 88.2% | 
| Jacobian determinant | - | 0.13% (folding) | 


| Performance metric | Without filtering | With filtering | 
|:-------------------|:------------:|:-----------:| 
| Dice coefficient | 98.08% | 99.59% | 
| Normalized cross correlation | 90.5% | 95.4% | 
| Jacobian determinant | 0.22%  | 0.25% (folding) | 


| Performance metric | Without filtering | With filtering | 
|:-------------------|:------------:|:-----------:| 
| Dice coefficient | 90.54% | 95.74% | 
| Normalized cross correlation | 81.9% | 88.2% | 
| Jacobian determinant | 0.40 | 0.13% (folding) | 



| Technical details | 
|:-------------------|:------------:| 
| MS image format | imzML, ibd | 
| Microscopic image format | tiff, png, jpg | 
| Availability | Open source at https://github.com/sslakkimsetty/msireg | 


# Session info

```{r session-info} 
sessionInfo()
```
