---
title: "Co-registration for the Harnblase TMA2 image"
author: "Sai Srikanth Lakkimsetty"
date: "Jan, 2023"
output:
    BiocStyle::html_document:
    toc: true 
vignette: >
  %\VignetteIndexEntry{Co-registration for the CRC dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, message=FALSE}
# A flag to manage chunk re-runs while knitting 
# (could also use cache option, but tricky) 
# Set SHARING to FALSE if you are re-knitting the code and variables are already 
# in the environment. Manage individual code chunk evaluation with "eval" option. 
SHARING <- FALSE 

if (SHARING) {
    library("msireg") 
} else {
    library("devtools")
    load_all()
} 

library("ggplot2")
RNGkind("L'Ecuyer-CMRG") 

setCardinalBPPARAM(MulticoreParam(stop.on.error=FALSE, progressbar=TRUE))
setCardinalVerbose(TRUE)
```

```{r style, echo=FALSE, results='asis', eval=TRUE}
BiocStyle::markdown()
```


```{r data-load, eval=SHARING} 
DATAPATH <- "/Users/lakkimsetty.s/Documents/9-data"
CURRPATH <- "10-Harnblase-TMAs"

opt_filename <- "HE_stained_TMA2.tiff"
opt <- EBImage::readImage(file.path(DATAPATH, CURRPATH, opt_filename))
opt <- flip(opt)
# opt <- resize(opt, w=dim(opt)[1]/2, h=dim(opt)[2]/2)

# mse_filename <- "TMA2.imzML"
# mse <- readMSIData(file.path(DATAPATH, CURRPATH, mse_filename))

mse_filename <- "TMA2"
mse <- readAnalyze(name=mse_filename,
    folder=file.path(DATAPATH, CURRPATH))
```


```{r data-load-2, eval=SHARING} 
# DATAPATH <- "/Users/lakkimsetty.s/Documents/9-data"
# CURRPATH <- "0-processed-in-R"
# load(file.path(DATAPATH, CURRPATH, "10-2-harnblase-TMA2.RData"))
```


```{r pre-proc, eval=SHARING}  
mse_mean <- summarizeFeatures(mse, FUN="mean", BPPARAM=MulticoreParam())

ref <- mse_mean |>
    peakPick(method="mad", SNR=2) |>
    peakAlign() |>
    peakFilter() |>
    process()

mse_peaks <- mse |>
    normalize(method="tic") |>
    peakBin(ref=mz(ref), tolerance=60, units="ppm") |>
    process()
``` 


```{r save-proc-imzml, eval=SHARING} 
# writeMSIData(mse_peaks,
#              file.path(DATAPATH,
#                        "0-processed-in-R/10-2-harnblase-TMA2.imzML"))
```


```{r mse-params, eval=SHARING} 
mse_attrs <- list(
    nX = dims(mse_peaks)[1], nY = dims(mse_peaks)[2],
    nF = dim(mse_peaks)[1], nP = dim(mse_peaks)[2],
    nXo = dim(opt)[1], nYo = dim(opt)[2]
)
```


```{r viz-ssc, eval=SHARING} 
set.seed(2)
# ssc <- spatialShrunkenCentroids(mse_peaks, r=c(0,1,2), s=c(0,3,6,9,12,15,18,21), k=c(5), 
#     method="gaussian", BPPARAM=MulticoreParam()) 
ssc <- spatialShrunkenCentroids(mse_peaks, r=c(1), s=c(0,3,6,9,12,15,18,21), k=c(5), 
    method="gaussian", BPPARAM=MulticoreParam()) 
# topf <- topFeatures(ssc, n=10, model=list(s=15))$mz 
topf <- topFeatures(ssc, n=100, sort.by=c("tstatistics"))$mz |> unique() 
image(mse_peaks, mz=topf, superpose=TRUE)
``` 


```{r viz-ssc-data}
message("Good features for initial visualization: ")
print(topf[1:8]) 

image(mse_peaks, mz=topf[2])
```


```{r rois, eval=SHARING} 
# RUN DIRECTLY FROM CONSOLE, SESSION WILL CRASH OTHERWISE 
mse_roi <- multiSelectROI(mse_peaks, mz=906.4585) 
opt_roi <- multiDrawROI(opt)
```


```{r opt-mask-img-crop, eval=SHARING} 
OUT <- cropToEdgesAndPadZeros(ref_mask=mse_roi, target_mask=opt_roi, 
                              target_img=opt) 
opt <- OUT$img 
opt_roi <- OUT$mask 
``` 

```{r coreg, eval=SHARING}
OUT <- coregister(mse_peaks, opt, mse_roi, opt_roi, spatial_scale=4, 
                  BPPARAM=MulticoreParam(), verbose=TRUE)
```


```{r, coreg-write-to-file, eval=SHARING} 
save.image("~/Documents/1-saved-envs/vignettes/10-2-harnblase-TMA2.RData") 
print(getwd())
WriteTransform(OUT$reg$TF, "/Users/lakkimsetty.s/Library/Mobile Documents/com~apple~CloudDocs/01-NU/phd/msireg/data/saved-transforms/10-2-harnblase-TMA2.hdf5")
```


```{r viz-coreg-prep, eval=TRUE} 
# library("devtools")
# load_all()
# load("~/Documents/1-saved-envs/vignettes/7-4-CRC-LM55.RData")
# OUT$reg$TF <- ReadTransform("data/saved-transforms/7-4-CRC-LM55.hdf5")
# out <- rlang::duplicate(OUT)
# out$fixed <- SimpleITK::as.image(out$msimg, isVector=FALSE)
# out$moving <- SimpleITK::as.image(out$opt, isVector=FALSE)

out$msimgx <- Resample(out$moving, out$reg$TF) |> as.array() |> Image() 

overlay <- overlayGridOnImage(out$msimgx)$overlay |> as.image()
grid <- overlayGridOnImage(out$msimgx)$grid |> as.image()

out$overlayx <- Resample(overlay, out$reg$TF) |> as.array() |> Image()
out$gridx <- Resample(grid, out$reg$TF) |> as.array() |> Image()
```


```{r viz-coreg} 
alpha <- 0.45
out$movingx <- Resample(out$moving, out$reg$TF)
.fixed <- out$fixed |> as.array() |> Image()
.moving <- out$moving |> as.array() |> Image()
.movingx <- out$movingx |> as.array() |> Image()

ol1 <- (.fixed * alpha + .moving * (1-alpha))
ol2 <- (.fixed * alpha + .movingx * (1-alpha))

def.par <- par(no.readonly=TRUE)

height <- 7
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

titles <- c("Optical image", "Mass spec. image",
            "Overlay before registration", "Overlay after registration")

par(mar=c(0.25, 0.25, 0.75, 0.25))
pl <- layout(mat=matrix(c(1:4), ncol=2, byrow=TRUE),
             heights=rep(height/2, 2), widths=rep(width/2, 2),
             respect=TRUE)

. <- sapply(c(1:4), function(x) {
    plot.new()
    rasterImage(list(.fixed, .moving, ol1, ol2)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE)
    title(titles[x])
})
```


```{r coreg-grid-overlay-viz} 
height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("warping on grid", "overlay with grid warping") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(out$gridx, out$overlayx)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


```{r metric-dice} 
spatial_scale <- 4 
DIM <- dim(out$MSIMG)[1:2] * spatial_scale
D_prereg <- diceCoefFromMasks(mse_roi, opt_roi) 
mse_roi_ss <- resizeAndPadImageToMatchDims(mse_roi, DIM) 
opt_roi_ss <- resizeAndPadImageToMatchDims(opt_roi, DIM) 
mse_roi_ss_x <- applyTransformOnMask(mse_roi_ss, out$reg$TF) 
D_postreg <- diceCoefFromMasks(mse_roi_ss_x, opt_roi_ss) 
```

Dice coefficient is computed from MSI and OPT masks. Dice coefficient before 
and after registration is `r D_prereg` and `r D_postreg` respectively. 

```{r metric-dice-viz} 
ol3 <- t(opt_roi_ss * 0.5 + mse_roi_ss * 0.5)
ol4 <- t(opt_roi_ss * 0.5 + mse_roi_ss_x * 0.5)

height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("Masks overlap before coreg", "Masks overlap after coreg") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(ol3, ol4)[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


```{r ncc} 
ncc_prereg <- normalizedCrossCorr(.fixed, .moving) 
ncc_postreg <- normalizedCrossCorr(.fixed, .movingx) 
```

Normalized cross correlation is a similarity metric that measure correlation 
between pixel intensities. It ranges from `-1` to `1`. The values before and 
after co-registration are `r ncc_prereg` and `r ncc_postreg` respectively. 


```{r jacobian} 
jac <- jacobianDeterminantFromTransform(out$reg$TF, ref_img=out$moving) 
hist(jac$jacobian_det, breaks=40)
```


```{r jacobian-viz-try1, eval=FALSE} 
jac$jacobian_det |> EBImage::display("raster") 
imagePlot(t(flip(jac$jacobian_det)))
```

```{r, jacobian-viz, eval=TRUE} 
def.par <- par(no.readonly=TRUE)
jacdet <- jac$jacobian_det
d <- dim(jacdet) 
asp <- d[2] / d[1] # h / w 
w <- 9 
h <- asp * w 

jacdet[jacdet <= 0] <- 0 
nfolds <- sum((jacdet == 0))
nshrinks <- length(unique(jacdet[(jacdet > 0) & (jacdet < 1)])) 
nequals <- sum(jacdet == 1)
nexpands <- length(unique(jacdet[(jacdet > 1)])) 

folds <- if (nfolds == 0) c() else c("#FF0000") # red 
shrinks <- hcl.colors(n=nshrinks, palette="Oranges") # white to oranges
equals <- if (nequals == 0) c() else c("#FFFFFF") # white
expands <- hcl.colors(n=nexpands, palette="Blues", rev=TRUE) # white to blues

layout(mat=matrix(1, nrow=1, ncol=1), heights=h, widths=w, respect=TRUE)
fields::imagePlot(flip(jacdet), asp=d[2]/d[1], xaxt="n", yaxt="n", 
          col=c(folds, shrinks, equals, expands), 
          lowerTriangle=TRUE)

par(def.par)
```


```{r, eval=FALSE} 
library("scales")
mesh <- expand.grid(x=c(1:d[1]), y=c(1:d[2])) 
mesh$z <- c(flip(jacdet)) 

ggplot(data=mesh, aes(x=x, y=y)) + 
    geom_tile(aes(fill=z)) + 
    scale_fill_gradient2(low = "red", 
                         mid = "white", 
                         high = "blue", 
                         midpoint=1) + 
    theme_void()
```


```{r, eval=FALSE}
def.par <- par(no.readonly=TRUE)
jacdet <- jac$jacobian_det
d <- dim(jacdet) 
asp <- d[2] / d[1] # h / w 
w <- 9 
h <- asp * w 

ncolors <- length(unique(jacdet))
layout(mat=matrix(1, nrow=1, ncol=1), heights=h, widths=w, respect=TRUE) 

x <- c(1:d[1]) 
y <- c(1:d[2])
image(x, y, (flip(jacdet)), zlim=range(jacdet), xlim=range(x), ylim=range(y), 
      xlab="", ylab="", xaxt="n", yaxt="n", 
      col=hcl.colors(ncolors))
```


The range of the determinant of the Jacobian is `r range(jac$jacobian_det)`. The 
histogram shows that most of the det values are around 1. That's good. The 
number of pixel locations where folding has occurred is 
`r sum(jac$jacobian_det <= 0)`.
Values 

* = 1 imply no area (or volume) change 
* (> 1) imply increased area
* (1, 0) imply reduced area 
* < 0 imply folding <- this is bad 


```{r gradient-field-cosine-dist, eval=SHARING} 
img1 <- .fixed
img2 <- .moving
img2x <- .movingx

grads1 <- matrixGradients(imageData(img1))
grads2 <- matrixGradients(imageData(img2))
grads2x <- matrixGradients(imageData(img2x))

.grads1 <- normalizeGradients(grads1)
.grads2 <- normalizeGradients(grads2)
.grads2x <- normalizeGradients(grads2x)

D <- dim(img1)

.grads <- array(c(.grads1, .grads2), dim=c(D, 4))
cos_dist <- apply(.grads, c(1,2), function(x) {
    if (identical(x[c(1:2)], x[c(3:4)])) return(1)

    (sum( x[c(1:2)] * x[c(3:4)] ))
}) |>
    array(dim=dim(img1))


.gradsx <- array(c(.grads1, .grads2x), dim=c(D, 4))
cos_distx <- apply(.gradsx, c(1,2), function(x) {
    if (identical(x[c(1:2)], x[c(3:4)])) return(1)

    (sum( x[c(1:2)] * x[c(3:4)] ))
}) |>
    array(dim=dim(img1))
```


```{r gradient-cosine-dist-plots, eval=TRUE} 
height <- 3.5
width <- 7 * (dim(.fixed)[1] / dim(.fixed)[2])

pl <- layout(mat=matrix(c(1:2), ncol=2, byrow=TRUE),
             heights=height, widths=rep(width/2, 2),
             respect=TRUE) 
titles <- c("Cosine dist before reg", "Cosine distance after reg") 
par(mar=c(0.25, 0.25, 0.75, 0.25))
. <- sapply(c(1:2), function(x) {
    plot.new()
    rasterImage(list(t(1-cos_dist), t(1-cos_distx))[[x]],
                xleft=0, xright=1, ytop=0, ybottom=1,
                interpolate=FALSE) 
    title(titles[x])
})
par(def.par)
```


# Session info

```{r session-info} 
sessionInfo()
```
